\LARGE \bf Planning with Attitude

### 2020-09-11 paper/iros_abstract.tex

    Many approaches have been taken to address the problem of motion planning and control
    on the $SE(3)$ manifold; most of these rely heavily on ideas and notation
    from differential geometry. Despite some impressive work in this field, these ideas
    have not been widely adopted and many continue to ignore the group structure of rotations,
    especially in the field of optimal control and motion planning. One part of this problem
    is that accounting for this structure requires low-level changes to the underlying
    optimization algorithms, which is difficult or impossible to implement when relying on
    commercial or even open-source optimization solvers.

    In the current abstract, we formulate a straightforward, generic method for adapting
    existing gradient-based algorithms to correctly account for the group structure of
    rotations. Based entirely on basic mathematical tools from linear algebra and
    calculus, our method is computationally efficient and should be both accessible and
    familiar to most roboticists. We then apply this method to our ALTRO solver
    \cite{howell2019altro} and demonstrate its performance on several constrained trajectory
    optimization problems.

@@

\section{Quaternion Differential Calculus} \label{sec:Quaternion_Calculus}
    Most modern methods for planning and control require derivatives with respect to the
    state vector. We present a simple but powerful method for taking derivatives of 
    quaternion-valued state vectors. The adaptation of this method to mixed state vectors,
    such as $SE(3)$, is straightforward.

@@

    \begin{figure}
        \centering
        \includegraphics[height=5cm]{figures/tangent_plane.tikz}
        \caption{
            When linearizing about a point $q$ on an sphere $\mathbb{S}^{n-1}$ in 
            n-dimensional space, the tangent space $T$ is a plane living in $\R^{n-1}$, 
            illustrated here with $n=3$. Therefore, when linearizing about a unit 
            quaternion $q \in \Q$, the space of differential rotations lives in $\R^3$.
        }
        \label{fig:tangent_plane}
    \end{figure}
    In practice we have found Rodrigues Parameters to be a very effective representation
    for three-dimensional differential rotations since they are computionally efficient
    and do not inherit the sign ambiguity of quaternions.

@@

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        unit quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(q \otimes \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
                        := \pdv{h}{q} G(q) 
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

@@

\section{Fast Constrained Trajectory Optimization on $SE(3)$}
    \subsection{Algorithmic Modifications}
    Here we briefly outline the changes to the ALTRO solver \cite{howell2019altro} to 
    efficiently solve trajectory optimization problems for systems with 3D rotations in the
    state vector. 
%     The AL-iLQR algorithm at the heart of ALTRO requires linearization of the
%     dynamics ($A \in \R^{n \times n}, B \in \R^{n \times m}$), the constraints ($\c_x \in 
%     \R^{p \times n}, c_u \in \R^{p \times m}$), and a quadratic expansion of the cost
%     function ($Q \in \R^{n \times n}, R \in \R^{m \times m}, q \in \R^n, r \in \R^m$). These
%     can be computed as follows:
    The linearization of the nonlinear discrete dynamics function $x_{k+1} = f(x_k,u_k)$ is 
    ``corrected'' using \ref{eq:quat_jacobian}:
    \begin{equation}
        A = G(f(x,u))^T \pdv{f}{x} G(x); \quad B = G(f(x,u))^T \pdv{f}{u}.
    \end{equation}
    We can similarly compute the linearization of a nonlinear constraint function $c(x,u)$
    using \ref{eq:quat_gradient}:
    \begin{equation}
        c_x = \pdv{c}{x} G(x); \quad c_u = \pdv{c}{x}.
    \end{equation}
    Lastly, we adjust the computation of the quadratic expansion of the augmented
    Lagrangian cost function $J(x,u) + \lambda^T c(x,u) + \half c(x,u)^T I_\mu c(x,u)$
    using \ref{eq:quat_gradient} and \ref{eq:quat_hessian}:
    \begin{subequations}
        \begin{align}
            Q &= & G(x)^T & \pdv[2]{J}{x} G(x) + c_x^T I_\mu c_x + I_3 \pdv{J}{x} x \\
            R &= &        & \pdv[2]{J}{u}      + c_u^T I_\mu c_u \\
            H &= &        & \pdv{J}{u}{x} G(x) + c_u^T I_\mu c_x \\
            q &= & G(x)^T & \pdv{J}{x} + c_x^T (I_\mu c(x,u) + \lambda) \\
            r &= &        & \pdv{J}{u} + c_u^T (I_\mu c(x,u) + \lambda).
        \end{align}
    \end{subequations}
    Here we define the \textit{state attitude Jacobian} $G(x)$ to be a block-diagonal
    matrix where the block is an identity matrix for any vector-valued state and $G(q)$ for
    any quaternion in the state vector. We refer the reader to the original ALTRO paper 
    \cite{howell2019altro} for futher details on the algorithm.

@@

    The algorithm only needs one small modification to the ``forward pass,'' where the
    trajectory is modified by simulating the system forward using the optimal feedback
    policy, $\delta u = K \delta x + d$, computed during the ``backward'' pass. Instead of 
    computing $\delta x$ using simple subtraction, we now compute the error state using the
    inverse Cayley map \eqref{eq:invcayley} for the quaternion states. The rest of the
    forward pass is unmodified.

@@

>>>>

    Many approaches \todo{add some citations here} have been taken to address the problem of motion planning and control
    on the $SE(3)$ group; most of these rely heavily on ideas and notation
    from differential geometry. Despite some impressive results, many of these ideas
    have not been widely adopted by the robotics community, and many practitioners continue to ignore the group structure of rotations. One issue preventing wider adoption is that accounting for this structure requires low-level changes to the underlying
    optimization algorithms, which is difficult or impossible when relying on existing
    off-the-shelf solvers.

    \todo{I'm worried that SE(3) and SO(3) results are getting confused/conflated here. I would explicitly talk about both, emphasize that we're handling SO(3) with the quaternion stuff, and that $SE(3) = SO(3) \times \mathbb{R}^3$ is a straightforward extension.}

    To address this lack of solver support, we formulate a straightforward, generic method for adapting
    existing Newton and gradient-based algorithms to correctly account for the group structure of
    rotations. Based entirely on basic mathematical tools from linear algebra and
    calculus, our method is computationally efficient and should be both accessible and
    familiar to most roboticists. We apply this method to the ALTRO solver
    \cite{howell2019altro} and demonstrate its performance on several constrained trajectory
    optimization problems.

@@

\section{Quaternion Differential Calculus} \label{sec:Quaternion_Calculus}
    Most modern methods for planning and control require derivatives with respect to the
    state vector. We present a simple but powerful method for taking derivatives of 
    quaternion-valued state vectors. The extension of this method to more general state vectors that contain Cartesian products of $SO(3)$, $SE(3)$, and $\mathbb{R}^n$ is straightforward.

@@

    \begin{figure}
        \centering
        \includegraphics[height=5cm]{figures/tangent_plane.tikz}
        \caption{
            When linearizing about a point $q$ on an sphere $\mathbb{S}^{n-1}$ in 
            n-dimensional space, the tangent space $T$ is a hyper-plane $\R^{n-1}$, 
            illustrated here with $n=3$. Therefore, when linearizing about a unit 
            quaternion $q \in \Q$, the space of differential rotations lives in $\R^3$.
        }
        \label{fig:tangent_plane}
    \end{figure}
    In practice, we have found Rodrigues Parameters to be a very effective representation
    for three-dimensional differential rotations since they are computionally efficient
    and do not inherit the sign ambiguity of quaternions.

@@

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        unit quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(q \otimes \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
                        := \pdv{h}{q} G(q) 
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is referred to as the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

@@

\section{Fast Constrained Trajectory Optimization on $SE(3)$}
    \subsection{Algorithmic Modifications}
    Here we briefly outline the changes to the ALTRO solver \cite{howell2019altro} to 
    efficiently solve trajectory optimization problems for systems with 3D rotations in the
    state vector. 
%     The AL-iLQR algorithm at the heart of ALTRO requires linearization of the
%     dynamics ($A \in \R^{n \times n}, B \in \R^{n \times m}$), the constraints ($\c_x \in 
%     \R^{p \times n}, c_u \in \R^{p \times m}$), and a quadratic expansion of the cost
%     function ($Q \in \R^{n \times n}, R \in \R^{m \times m}, q \in \R^n, r \in \R^m$). These
%     can be computed as follows:
    The linearization of the nonlinear discrete dynamics function $x_{k+1} = f(x_k,u_k)$ is 
    ``corrected'' using \ref{eq:quat_jacobian}:
    \begin{equation}
        A = G(f(x,u))^T \pdv{f}{x} G(x); \quad B = G(f(x,u))^T \pdv{f}{u}.
    \end{equation}
    We can similarly compute the linearization of a nonlinear constraint function $c(x,u)$
    using \ref{eq:quat_gradient}:
    \begin{equation}
        c_x = \pdv{c}{x} G(x); \quad c_u = \pdv{c}{x}.
    \end{equation}
    Lastly, we adjust the computation of the quadratic expansion of the augmented
    Lagrangian \todo{I would leave out the AL stuff and just talk about a generic cost function here to make it more general} cost function $J(x,u) + \lambda^T c(x,u) + \half c(x,u)^T I_\mu c(x,u)$
    using \ref{eq:quat_gradient} and \ref{eq:quat_hessian}:
    \begin{subequations}
        \begin{align}
            Q &= & G(x)^T & \pdv[2]{J}{x} G(x) + c_x^T I_\mu c_x + I_3 \pdv{J}{x} x \\
            R &= &        & \pdv[2]{J}{u}      + c_u^T I_\mu c_u \\
            H &= &        & \pdv{J}{u}{x} G(x) + c_u^T I_\mu c_x \\
            q &= & G(x)^T & \pdv{J}{x} + c_x^T (I_\mu c(x,u) + \lambda) \\
            r &= &        & \pdv{J}{u} + c_u^T (I_\mu c(x,u) + \lambda).
        \end{align}
    \end{subequations}
    Here we define the \textit{state attitude Jacobian} $G(x)$ to be a block-diagonal
    matrix where the block is an identity matrix for any vector-valued state and $G(q)$ for
    any quaternion in the state vector. We refer the reader to the original ALTRO paper 
    \cite{howell2019altro} for futher details on the algorithm.

@@

    The algorithm only needs one small modification to the ``forward pass,'' where the
    trajectory is modified by simulating the system forward using the optimal feedback
    policy, $\delta u = K \delta x + d$, computed during the ``backward'' pass. Instead of 
    computing $\delta x$ using simple subtraction, we now compute the error state using the
    inverse Cayley map \eqref{eq:invcayley} for the quaternion states. The rest of the
    forward pass is unmodified. \todo{I think you can safely omit some of these details and just refer to the ALTRO paper.}

@@

        \todo{It would be great to compare to both the minimal iLQR stuff and SNOPT or IPOPT with quaternion norm constraints + slack variables (referred to as the ``embedding approach'').}
